A cautionary note on the use of Ornstein Uhlenbeck models in macroevolutionary studies

### 2015-05-15 Manuscript/ohyou.tex
dealt with  CV and AM's comments plus some additions

  The OU model is most commonly used to model the evolution of a single continuous character (Table \ref{table.uses}; Supporting Information). 
  Usually several models of evolution \citep[e.g., Brownian motion, OU, Early burst etc.;][]{Harmon:2010aa,cooper2010body} are fit to the same continuous character and model selection is then used to determine which model best fits the data.  
  OU models can be fit with one optimal trait value, or multiple different optima \citep{Butler:2004aa,beaulieu2012modeling}. 
  The latter represents evolution under multiple selective regimes, and may be more biologically realistic for many datasets. 
  OU models with various numbers of optima are often included in the pool of evolutionary models being compared \citep[e.g.,][]{christin2013anatomical}(Table \ref{table.uses})
  OU models are also commonly used to control for phylogeny in correlative analyses (Table \ref{table.uses}; Supporting Information).  
  Phylogenetic generalised least squares (PGLS) models incorporate information about the relationships among species into the error term of a generalised least squares model. 
  This error term generally consists of a variance-covariance matrix of the phylogeny, but various transformations are used (e.g., Pagel's $\lambda$; \citealp{Pagel:1997aa}) to improve the fit of the model to the data. 
  The $\alpha$ parameter from an OU model can also be used to transform the tree and the variance-covariance matrix.
  This is rarely interpreted as corresponding to any kind of process,
  ; instead it improves the fit of the PGLS models \citep[e.g.,][]{blankers2012ecological}. 
  It is not clear that it was originally intended that the OU model would be used in such a context.
  Finally, the OU model is also used to reconstruct ancestral states \citep{martins1999estimation} and to detect clade-wide convergent evolution \citep{ingram2013surface}.

@@

    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large, all imprint of history is lost and the trait evolution is essentially a rapid burst at the present.

@@

	    We (AM and CV) repeated the Maximum Likelihood simulations in a Bayesian framework implemented in BayesTraits \citep{pagel2013bayestraits}. 
      We determined fit to an OU model using Bayes factors estimated from a stepping stone sampling procedure \citep{xie2010improving}. 
      The marginal likelihoods of the models were calculated using a stepping stone sampler in which fifty stones were drawn from a beta distribution with (alpha = 0.4 and beta = 1). 
      Each stone was sampled for 20000 iterations (with the first 5000 iterations discarded). 
      We treated Bayes factors \textgreater 2 as evidence favouring the OU model. We ran the MCMC chains for $1x10^6$ iterations, disregarding the first $1x10^4$ as burn-in.
      Following burn-in the chains were sampled every 1000 to ensure independence of each consecutive sample. 
      Multiple independent chains were run for each analysis to ensure convergence was reached. 
      An important challenge for Bayesian approaches is selection of appropriate priors. We used three alternative sets of priors on $\alpha$: (i) an exponential distribution with mean = 1; (ii), an exponential distribution with mean = 10, and (iii) a uniform distribution bounded at 0 and 20. 
      For all analyses we used a uniform -100 to 100 prior for $\mu$ and uniform 0 to 100 prior for $\sigma^2$.

      Table \ref{tables:bayesian} shows the results from the exponential prior with mean = 10. This is a broad, liberal prior, but our results are similar regardless of priors. %add links to this
      The Bayesian approach results in highly conservative rejection rates regardless for all tree shapes in the absence of measurement error. 
      While this is encouraging from the perspective of falsely rejecting the Brownian null model, it is also indicative of potentially low statistical power. 
      The Bayesian approach also appears to more readily handle low levels of measurement error (Table \ref{tables:bayesian_error}). 
      With 1\% measurement error the Bayesian approach retains acceptable rejection rates (\textless 0.05) for trees of up to 150 tips. 
      However, more error or larger trees result in the frequent rejection of the Brownian model. 
      As noted above, this is not an issue of Type I and it is entirely correct that the Brownian model is rejected. 
      In such cases emphasis should shift to how the $\alpha$ parameter is interpreted. 
      At this stage we are cautious in recommending Bayesian methods for fitting OU models but instead suggest that this may be a fruitful area for further investigation. 
      See \citet{pennell2015model} for more details on using Bayesian methods in evolutionary models.
    \end{enumerate}

@@

  \subsection{Recommendations for interpreting $\alpha$}    
    \begin{enumerate}[(i)]
    \item \textbf{Consider plausible alternative hypotheses.}
    Because the OU model was proposed with evolutionary process in mind, alternative and arguably more parsimonious explanations for favouring OU models and interpreting non-zero $\alpha$ are often overlooked. 
    The effects of measurement error in particular suggest that $\alpha$ must always be interpreted with caution. 
    Many issues of misinterpretation can be solved by carefully inspecting the $\alpha$ parameter when an OU model is favoured. 
    Often, when Likelihood ratio tests suggest the OU model should be favoured over the Brownian model, estimates of $\alpha$ are actually very close to zero, i.e. very close to Brownian.
    In these circumstances, it is likely that measurement error, or similar, is being mopped up by the extra parameters in the OU model. 
    Thus this does not reflect any kind of OU process underlying the data. 
    The similarity between Brownian and OU models with small $\alpha$ is demonstrated in (Figure \ref{figure:ouhistory}). 
    At the other extreme, as values of $\alpha$ become larger, the effects of changing $\alpha$ on model predictions are increasingly small and large values of $\alpha$ are indistinguishable from white-noise.

@@

    \noindent
    If $t_\frac{1}{2}$ is short relative to the branch lengths of the phylogeny, evolution towards the optimum trait value is fast, residual phylogenetic correlations are weak, and there is little influence of the past on trait values \citep{hansen1997stabilizing}.
    We would not advise interpreting $t_\frac{1}{2}$ as literally being ``the time it takes for a species entering a new niche to evolve halfway toward its new expected optimum''.
    However, if $t_\frac{1}{2}$ is extremely large, it suggests that if an OU process is acting, it is extremely weak, thus should not be interpreted as evidence of any kind of process. 
    As a further note of caution, it is important to recognise that biases in the estimation of $\alpha$ would lead to similar biases in $t_\frac{1}{2}$.

    \item \textbf{If possible, include ancillary data.}
    If data on fossils are available then these could be incorporated into the analysis \citep{Slater:2012ab}.
    This is implemented in GEIGER \citep{Harmon:2008aa}.
    A caution here is that the OU model for non-ultrametric trees has to be carefully parameterized because for non-ultrametric trees the co-variances depend on both the shared distances between species and the distance of a node to the nearest tip (see equation \ref{equation:OUcov} in the Appendix). 
    This creates potential problems in parameterization and in interpretation because the variance-covariance matrix is no longer tree-like \citep{slater2014correction}.
    Some current implementations of the OU model are based on transforming the tree directly, rather than transforming the variance co-variance matrix \citep[e.g., MOTMOT;][]{Thomas:2011aa}. 
    These implementations should not be used with fossil data.
   \end{enumerate}

\section{Conclusions and outstanding issues}
  A recurring theme from our simulations is that interpretation of OU models is not straightforward. More focus is needed on the interpretation of $\alpha$ rather than simply model fit. Even when the OU model is favoured, $\alpha$ may be so small as to be indistinguishable from Brownian motion in any biological sense. It would clearly be very useful to have estimates of measurement error for all species traits, however the inclusion of species-specific variances has to be done carefully \citep[e.g.,][]{grafen1989phylogenetic}. Several approaches for accounting for error have been proposed and warrant further study for OU models and more widely for models of trait evolution. %check references.

@@

\section{Epilogue: (vaguely) reproducible science}
  At the symposium that generated this special issue, one of us (NC) gave a talk on reproducibility in science. 
  We have therefore tried to make this paper as reproducible as possible. 
  All simulated phylogenies, data, and R code are available on GitHub: \href{https://github.com/nhcooper123/OhYou}{github.com/nhcooper123/OhYou}.
  However, although our R code is provided, it is messy and difficult to use. 
  We also do not provide an automated way of reproducing the data collection for our literature review.
  Nor do we use tools like Travis CI (travis-ci.org), Docker (www.docker.com), or packrat \citep{Ushey:2015aa} etc. to increase the reproducbility of our analyses.
  Additionally, we use BayesTraits to run our Bayesian analyses. 
  Although free to download, this software is not an Open Source, which has implications for reproducibility.
  We included this epilogue to highlight that reproducibility is really hard, but we are trying to improve! 
  With a little effort most people should be able to produce something vaguely reproducible, and move us all slightly closer to truly reproducible science.

\section{Acknowledgments}
  Thanks to Mark Pagel, Matt Pennell, Graham Slater and Rich FitzJohn for fruitful discussions about OU models, and three anonymous reviewers for helpful comments on a previous version of this paper. NC was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696. GHT was supported by a Royal Society University Research Fellowship. AM was supported by BBSRC grant BB/K004344/1 and the computing time was funded by European Research Council Grant no. 268744, Mother Tongue.

>>>>

  The OU model is most commonly used to model the evolution of a single continuous character (Table \ref{table.uses}; Supporting Information). 
  Usually several models of evolution \citep[e.g., Brownian motion, OU, Early burst etc.;][]{Harmon:2010aa,cooper2010body,cardillo2015geographic,slater2015iterative} are fit to the same continuous character and model selection is then used to determine which model best fits the data.  
  OU models can be fit with one optimal trait value, or multiple different optima \citep{Butler:2004aa,beaulieu2012modeling}. 
  The latter represents evolution under multiple selective regimes, and may be more biologically realistic for many datasets. 
  OU models with various numbers of optima are often included in the pool of evolutionary models being compared \citep[e.g.,][]{christin2013anatomical}(Table \ref{table.uses})
  OU models are also commonly used to control for phylogeny in correlative analyses (Table \ref{table.uses}; Supporting Information).  
  Phylogenetic generalised least squares (PGLS) models incorporate information about the relationships among species into the error term of a generalised least squares model. 
  This error term generally consists of a variance-covariance matrix of the phylogeny, but various transformations are used (e.g., Pagel's $\lambda$; \citealp{Pagel:1997aa}) to improve the fit of the model to the data. 
  The $\alpha$ parameter from an OU model can also be used to transform the tree and the variance-covariance matrix.
  This is rarely interpreted as corresponding to any kind of process,
  ; instead it improves the fit of the PGLS models \citep[e.g.,][]{blankers2012ecological}. 
  It is not clear that it was originally intended that the OU model would be used in such a context.
  Finally, the OU model is also used to reconstruct ancestral states \citep{martins1999estimation} and to detect clade-wide convergent evolution \citep{ingram2013surface}.

@@

    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large, all imprint of history is lost and the trait evolution is essentially a rapid burst at the present.
    Note that $\alpha$ scales with tree height, so it needs to be interpreted relative to this. 
    Generally the simplest solution is to rescale tree heights to 1 \citep[e.g.,][simulations below]{ives2010phylogenetic}. 
    \citet{ives2010phylogenetic} suggest interpreting $log( \alpha)$ (after rescaling trees to a height of 1) rather than raw $\alpha$. They equate $-log( \alpha) = -4$ as a very low, almost Brownian value, and $-log( \alpha) = 4$ as a very high value. 
    Others \citep[e.g.][]{slater2015iterative,slater2013robust,hansen2012interpreting} prefer to use the ``phylogenetic half-life'': $t_\frac{1}{2}$ (see ``Recommendations for interpreting $\alpha$'' below).

@@

	    We (AM and CV) repeated the Maximum Likelihood simulations in a Bayesian framework implemented in BayesTraits \citep{pagel2013bayestraits}. 
      We determined fit to an OU model using Bayes factors estimated from a stepping stone sampling procedure \citep{xie2010improving}. 
      The marginal likelihoods of the models were calculated using a stepping stone sampler in which fifty stones were drawn from a beta distribution with (alpha = 0.4 and beta = 1). 
      Each stone was sampled for 20000 iterations (with the first 5000 iterations discarded). 
      We treated Bayes factors \textgreater 2 as evidence favouring the OU model \citep{kass1995bayes}. 
      We ran the MCMC chains for $1x10^6$ iterations, disregarding the first $1x10^4$ as burn-in.
      Following burn-in the chains were sampled every 1000 to ensure independence of each consecutive sample. 
      Multiple independent chains were run for each analysis to ensure convergence was reached. 
      An important challenge for Bayesian approaches is selection of appropriate priors. We used three alternative sets of priors on $\alpha$: (i) an exponential distribution with mean = 1; (ii), an exponential distribution with mean = 10, and (iii) a uniform distribution bounded at 0 and 20. 
      For all analyses we used a uniform -100 to 100 prior for $\mu$ and uniform 0 to 100 prior for $\sigma^2$.

      Table \ref{tables:bayesian} shows the results from the exponential prior with mean = 10. This is a broad, liberal prior, but our results are similar regardless of priors. %add links to this
      The Bayesian approach results in highly conservative rejection rates regardless for all tree shapes in the absence of measurement error. 
      While this is encouraging from the perspective of falsely rejecting the Brownian null model, it is also indicative of potentially low statistical power, although testing would be required to confirm this. 
      The Bayesian approach also appears to more readily handle low levels of measurement error (Table \ref{tables:bayesian_error}). 
      With 1\% measurement error the Bayesian approach retains acceptable rejection rates (\textless 0.05) for trees of up to 150 tips. 
      However, more error or larger trees result in the frequent rejection of the Brownian model. 
      As noted above, this is not an issue of Type I and it is entirely correct that the Brownian model is rejected. 
      In such cases emphasis should shift to how the $\alpha$ parameter is interpreted. 
      At this stage we are cautious in recommending Bayesian methods for fitting OU models but instead suggest that this may be a fruitful area for further investigation. 
      See \citet{pennell2015model} for more details on using Bayesian methods in evolutionary models.
    \end{enumerate}

@@

  \subsection{Recommendations for interpreting $\alpha$}    
    \begin{enumerate}[(i)]
    \item \textbf{Consider plausible alternative hypotheses.}
    Because the OU model was proposed with evolutionary process in mind, alternative and arguably more parsimonious explanations for favouring OU models and interpreting non-zero $\alpha$ are often overlooked. 
    The effects of measurement error in particular suggest that $\alpha$ must always be interpreted with caution. 
    Many issues of misinterpretation can be solved by carefully inspecting the $\alpha$ parameter when an OU model is favoured. 
    Often, when Likelihood ratio tests suggest the OU model should be favoured over the Brownian model, estimates of $\alpha$ are actually very close to zero (e.g., examples in \citealp{Harmon:2010aa}), i.e. very close to Brownian.
    In these circumstances, it is likely that measurement error, or similar, is being mopped up by the extra parameters in the OU model. 
    Thus this does not reflect any kind of OU process underlying the data. 
    The similarity between Brownian and OU models with small $\alpha$ is demonstrated in (Figure \ref{figure:ouhistory}). 
    At the other extreme, as values of $\alpha$ become larger, the effects of changing $\alpha$ on model predictions are increasingly small and large values of $\alpha$ are indistinguishable from white-noise.

@@

    \noindent
    If $t_\frac{1}{2}$ is short relative to the branch lengths of the phylogeny, evolution towards the optimum trait value is fast, residual phylogenetic correlations are weak, and there is little influence of the past on trait values \citep{hansen1997stabilizing}.
    $t_\frac{1}{2}$ equal to the height of the phylogeny is a moderate value \citep{hansen2012interpreting}.
    We would not advise interpreting $t_\frac{1}{2}$ as literally being ``the time it takes for a species entering a new niche to evolve halfway toward its new expected optimum''.
    However, if $t_\frac{1}{2}$ is extremely large, it suggests that if an OU process is acting, it is extremely weak, thus should not be interpreted as evidence of any kind of process. 
    As a further note of caution, it is important to recognise that biases in the estimation of $\alpha$ would lead to similar biases in $t_\frac{1}{2}$.

    \item \textbf{If possible, include ancillary data.}
    If data on fossils are available then these could be incorporated into the analysis \citep{Slater:2012ab}, provided their placement is well known.
    This is implemented in GEIGER \citep{Harmon:2008aa}.
    A caution here is that the OU model for non-ultrametric trees has to be carefully parameterized because for non-ultrametric trees the co-variances depend on both the shared distances between species and the distance of a node to the nearest tip (see equation \ref{equation:OUcov} in the Appendix). 
    This creates potential problems in parameterization and in interpretation because the variance-covariance matrix is no longer tree-like \citep{slater2014correction}.
    Some current implementations of the OU model are based on transforming the tree directly, rather than transforming the variance co-variance matrix \citep[e.g., MOTMOT;][]{Thomas:2011aa}. 
    These implementations should not be used with fossil data.
   \end{enumerate}

\section{Conclusions and outstanding issues}
  A recurring theme from our simulations is that interpretation of OU models is not straightforward. More focus is needed on the interpretation of $\alpha$ rather than simply model fit. Even when the OU model is favoured, $\alpha$ may be so small as to be indistinguishable from Brownian motion in any biological sense. It would clearly be very useful to have estimates of measurement error for all species traits, however the inclusion of species-specific variances has to be done carefully \citep[e.g.,][]{grafen1989phylogenetic}. Several approaches for accounting for error have been proposed and warrant wider implementation for OU models and other models of trait evolution \citep[e.g.,][]{lynch1991methods,hansen2012interpreting,martins1997phylogenies,rohlfs2013modeling,ives2007within}, however it is outside our scope to explore these here.

@@

\section{Epilogue: Open and (vaguely) reproducible science}
  At the symposium that generated this special issue, one of us (NC) gave a talk on Open Science and reproducibility. 
  We have therefore tried to make this paper as open and reproducible as possible. 
  All simulated phylogenies, data, and R code are available on GitHub: \href{https://github.com/nhcooper123/OhYou}{github.com/nhcooper123/OhYou}.
  However, although our R code is provided, it is messy and difficult to use. 
  We also do not provide an automated way of reproducing the data collection for our literature review.
  Nor do we use tools like Travis CI (travis-ci.org), Docker (www.docker.com), or packrat \citep{Ushey:2015aa} etc. to increase the reproducbility of our analyses.
  Additionally, we use BayesTraits to run our Bayesian analyses. 
  Although free to download, this software is not an Open Source, which has implications for long-term development of the methods.
  We included this epilogue to highlight that reproducibility is really hard, but we are trying to improve! 
  With a little effort most people should be able to produce something vaguely reproducible, and to provide their data and code, moving us all slightly closer to truly reproducible and Open Science.

\section{Acknowledgments}
  Thanks to Mark Pagel, Matt Pennell, Graham Slater and Rich FitzJohn for fruitful discussions about OU models, and three anonymous reviewers for helpful comments on a previous version of this paper. NC was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696. GHT was supported by a Royal Society University Research Fellowship. CV was supported by a Leverhulme Trust Research Project Grant RPG-2013-185. AM was supported by BBSRC grant BB/K004344/1 and the computing time was funded by European Research Council Grant no. 268744, ``MotherTongue''.
